Non-Vertical Cultural Transmission, Assortment, \\and the Evolution of Cooperation

### 2020-10-25 ms/main.tex
refactored analysis to appendices

To determine the equilibria in our model, we analyze the fixed points of \autoref{eq:nextgen_juveniles}.
We define $f(\hat{p}) = \bar{w}(\hat{p}' - \hat{p})$.
Then, using \emph{SymPy} \citep{Meurer2017}, a Python library for symbolic mathematics, this simplifies to
\begin{equation} \label{eq:general_case_polynomial}
\begin{aligned}
  f(\hat{p}) &= \bar{w}(\hat{p}'-\hat{p}) =
  \beta_1 \hat{p}^3 + \beta_2 \hat{p}^2 + \beta_3 \hat{p} \;,
\end{aligned}
\end{equation}
where 
\begin{equation} \label{eq:polynomial_coefficients}
\begin{aligned}
  \beta_1 &= \big[c(1-v) - b (1-\alpha v)\big] (T_A-T_B) \;, \\
  \beta_2 &= -\beta_1 -\beta_3 \;, \\
  \beta_3 &= \alpha bvT_A - cv(1-T_B) + (T_A-T_B) \;.
\end{aligned}
\end{equation}

In the general case where $T_A \neq T_B$, the coefficient $\beta_1$ is not necessarily zero, and $f(\hat{p})$ is a cubic polynomial.
Therefore, three equilibria may exist, two of which are
$\hat{p} = 0 $ and $\hat{p} = 1$, and the third is
\begin{equation} \label{eq:general_equilibrium}
  \hat{p}^* =  
  \frac{\beta_3}{\beta_1} =
  \frac{\alpha bvT_A - cv(1-T_B) + (T_A-T_B)}{\big[c(1-v) - b (1-\alpha v)\big] (T_A-T_B)} \;.
\end{equation}

To present this analysis in terms of conditions on the cost of cooperation $c$, we define the cost boundaries, $\gamma_1$ and $\gamma_2$, and the vertical transmission threshold, $\hat v$,

\begin{equation} \label{eq:cost_boundaries_v_threshold} 
\gamma_1 = \frac{b v \alpha T_A + (T_A - T_B)}{v(1-T_B)}, \quad
\gamma_2 = \frac{b v \alpha T_B + (1+b) (T_A - T_B)}{v(1-T_B) + (1-v)(T_A-T_B)}, \quad
\hat v = \frac{T_B - T_A}{1-T_A} \;.
\end{equation}

First, assume $T_A<T_B$.
$\beta_3<0$ requires $\gamma_1<c$.
For $\beta_3<\beta_1$ we need $c\big[v(1-T_B) + (1-v)(T_A-T_B)\big] > bv\alpha T_B + (1+b)(T_A-T_B)$.
Note that the expression in the square brackets is positive if and only if $v > \hat v$.
Thus, for $\beta_3<\beta_1$ we need $v > \hat v$ and $\gamma_2 < c$ or $v < \hat v$ and $c < \gamma_2$,
and for $0<\beta_3<\beta_1$ we need $v > \hat v$ and $\gamma_2 < c < \gamma_1$, or $v < \hat v$ and $c < \min(\gamma_1, \gamma_2)$. For $\beta_1<\beta_3$ we need $v > \hat v$ and $c<\gamma_2$ or $v < \hat v$ and $\gamma_2<c$.
However, some of these conditions cannot be met, since $v < \hat v$ implies $c<1<\gamma_2$.

The following result summarizes the possible outcomes of the model (\autoref{eq:nextgen_juveniles}).
\\

@@

% in beta
%\begin{enumerate}
%\item \emph{Fixation of cooperation}: 
%	if \emph{(i)} $T=T_A=T_B$ and $c < b\cdot \frac{\alpha T}{1-T}$; or
%	if \emph{(ii)} $T_A>T_B$ and $0<\beta_3$; or 
%	if \emph{(iii)} $T_A<T_B$ and $\beta_1<\beta_3$.
%\item \emph{Fixation of the defection}: 
%	if \emph{(iv)}  $T=T_A=T_B$ and $c > b\cdot \frac{\alpha T}{1-T}$; or 
%	if \emph{(v)} $T_A>T_B$ and $\beta_3<\beta_1<0$; or 
%	if \emph{(vi)} $T_A<T_B$ and $\beta_3<0$.
%\item \emph{Coexistence of both phenotypes at $\hat{p}^*$}: 
%	if \emph{(vii)} $T_A < T_B$ and $0<\beta_3<\beta_1$.
%\item \emph{Fixation of either phenotype depending on initial frequency}:
%	if \emph{(viii)}  $T_A>T_B$ and $\beta_1<\beta_3<0$.
%\end{enumerate}

@@

We now analyze the evolution of social association under perfect vertical transmission, $v=1$.
We assume a population at a stable coexistence of the two phenotypes, cooperation $A$, and defection $B$, such that the frequency of $A$ among parents is $\tilde{p}^*$ (\autoref{eq:general_equilibrium}).
Note that for stable coexistence to occur, we assume horizontal bias for defection, $T_A < T_B$, and an intermediate cost of cooperation, $\gamma_2 < c < \gamma_1$ (\autoref{eq:cost_boundaries_v_threshold})

The stable population mean fitness is $\bar{w}^* = 1 + \tilde{p}^*(b-c)$, which is increasing in $\tilde{p}^*$. 
Therefore, to determine the effect of increasing $\alpha$ on the stable population mean fitness, we must analyze its effect on $\tilde{p}^*$, 
\begin{equation} \label{eq:mean_fitness_stable_coexistence_derivative}
  \frac{\partial \tilde{p}^*}{\partial \alpha} 
%  =\frac{-b^2T_A (1-\alpha) (T_A-T_B) + b(T_A-T_B)(c(1-T_B) - b \alpha T_A - (T_A-T_B))}{b^2 (1-\alpha)^2 (T_A-T_B)^2} 
%  = \frac{b^2T_A(T_B-T_A)-bc(1-T_B)(T_B-T_A) - b(T_B-T_A)^2}{b^2 (1-\alpha)^2 (T_A-T_B)^2} \;.
  = \frac{b T_A - c(1-T_B) + (T_A-T_B)}{b (1-\alpha)^2 (T_B-T_A)} \;.
\end{equation} 
Note that stable coexistence implies $c<\gamma_1$, and because $\alpha<1$, we have
\begin{equation}
c < \gamma_1 = \frac{b \alpha T_A + (T_A-T_B)}{1-T_B} < \frac{b T_A + (T_A-T_B)}{1-T_B}.
\end{equation} 
Therefore, the numerator in \autoref{eq:mean_fitness_stable_coexistence_derivative} is positive.
Since $T_A<T_B$, the denominator in \autoref{eq:mean_fitness_stable_coexistence_derivative} is also positive, and hence the derivative $\partial \tilde{p}^* / \partial \alpha$ is positive.
Thus, the population mean fitness increases as social association $\alpha$ increases.

We found that if social association increases, the population mean fitness also increases.
But can an allele that increases social association evolve? 
To answer this question, we extend our model to include a ``modifier locus''~\citep{Liberman1986,Liberman1988} that determines social association, but has no direct effect on fitness.
We assume that the modifier locus has two alleles, $M$ and $m$, which induce social associations $\alpha_1$ and $\alpha_2$, respectively.
Suppose that the population evolves to a stable equilibrium when only allele $M$ is present; we check the local stability of this equilibrium to the introduction of allele $m$ at the modifier locus; This is called ``external stability'', see \citet{Altenberg2017}.

We define the frequencies of the pheno-genotypes $AM$, $BM$, $Am$, and $Bm$ as $\vec{\tilde{p}}= (\tilde{p}_1$, $\tilde{p}_2$, $\tilde{p}_3$, $\tilde{p}_4)$. 
The frequencies of the pheno-genotypes in the next generation are defined by the recursion system 
\begin{equation} \label{eq:next_gen_p_1}
  \begin{aligned}
	% p1
  \bar{w}\tilde{p}_1' & = \tilde{p}_1(\tilde{p}_1+\tilde{p}_3)(1+b-c)(1 - (1-\alpha_1)(\tilde{p}_2 + \tilde{p}_4)T_B) \\
  &+ \tilde{p}_1(\tilde{p}_2 + \tilde{p}_4)(1-c)(1-\alpha_1T_B(\tilde{p}_1 + \tilde{p}_3) - T_B(\tilde{p}_2 + \tilde{p}_4)) \\
  &+ \tilde{p}_2(\tilde{p}_1 + \tilde{p}_3)(1+b)T_A(\tilde{p}_1 + \tilde{p}_3 + \alpha_1(\tilde{p}_2 + \tilde{p}_4)) \\
  &+ \tilde{p}_2(\tilde{p}_2 + \tilde{p}_4)(\tilde{p}_1 + \tilde{p}_3)(1-\alpha_1)T_A \;,
\\
	% p2
  \bar{w}\tilde{p}_2' & = \tilde{p}_2(\tilde{p}_2+\tilde{p}_4)(1 - (1-\alpha_1)(\tilde{p}_1 + \tilde{p}_3)T_A) \\
  &+ \tilde{p}_2(\tilde{p}_1 + \tilde{p}_3)(1+b)(1-\alpha_1 T_A(\tilde{p}_2 + \tilde{p}_4) - T_A(\tilde{p}_1 + \tilde{p}_3))\\
  &+ \tilde{p}_1(\tilde{p}_2 + \tilde{p}_4)(1-c)(\alpha_1 T_B + (1-\alpha_1)(\tilde{p}_2 + \tilde{p}_4)T_B) \\
  &+ \tilde{p}_1(\tilde{p}_1 + \tilde{p}_3)(1+b-c)(1-\alpha_1)(\tilde{p}_2 + \tilde{p}_4)T_B \;, 
\\
	% p3
  \bar{w}\tilde{p}_3' & = \tilde{p}_3(\tilde{p}_1+\tilde{p}_3)(1+b-c)(1 - (1-\alpha_2)(\tilde{p}_2 + \tilde{p}_4)T_B) \\
  &+ \tilde{p}_3(\tilde{p}_2 + \tilde{p}_4)(1-c)(1-\alpha_2T_B(\tilde{p}_1 + \tilde{p}_3) - T_B(\tilde{p}_2 + \tilde{p}_4)) \\
  &+ \tilde{p}_4(\tilde{p}_1 + \tilde{p}_3)(1+b)T_A(\tilde{p}_1 + \tilde{p}_3 + \alpha_2(\tilde{p}_2 + \tilde{p}_4)) \\
  &+ \tilde{p}_4(\tilde{p}_2 + \tilde{p}_4)(\tilde{p}_1 + \tilde{p}_3)(1-\alpha_2)T_A \;, 
\\
	% p4
  \bar{w}\tilde{p}_4' & = \tilde{p}_4(\tilde{p}_2 + \tilde{p}_4)(1 - (1-\alpha_2)(\tilde{p}_1 + \tilde{p}_3)T_A) \\
  &+ \tilde{p}_4(\tilde{p}_1 + \tilde{p}_3)(1+b)(1-\alpha_2T_A(\tilde{p}_2 + \tilde{p}_4)-T_A(\tilde{p}_1 + \tilde{p}_3))\\
  &+ \tilde{p}_3(\tilde{p}_2 + \tilde{p}_4)(1-c)(\alpha_2T_B + (1-\alpha_2)(\tilde{p}_2 + \tilde{p}_4)T_B) \\
  &+ \tilde{p}_3(\tilde{p}_1 + \tilde{p}_3)(1+b-c)(1-\alpha_2)(\tilde{p}_2 + \tilde{p}_4)T_B \;,
  \end{aligned}
\end{equation}
%which we can write as 
%\begin{equation}
%\bar{w} \vec{\tilde{p}} = F(\vec{\tilde{p}}),
%\end{equation}
where $\bar{w} = 1 + (b-c)(\tilde{p}_1 + \tilde{p}_3)$ is the population mean fitness.% and $\vec{\tilde{p}} = (\tilde{p}_1,\tilde{p}_2,\tilde{p}_3,\tilde{p}_4)$.

If both eigenvalues of $\mathbf{L}^*_{ex}$ are less than 1, then the equilibrium $\vec{\tilde{p}^*}$ is externally stable and allele $m$ cannot invade the population of allele $M$. 
These eigenvalues are the root of the characteristic polynomial $\lambda^2 -\lambda (X+Q) + XQ - YZ$; the leading eigenvalue $\lambda_1$ is the largest of them,
\begin{equation} \label{eq:eigenvalues}
  \lambda_1 = \frac{X+Q + \sqrt{(X+Q)^2 - 4(XQ-YZ)}}{2} \;.
\end{equation}
Using \emph{SymPy} \citep{Meurer2017}, a Python library for symbolic mathematics, we find that $\lambda_1 > 1$ if 
\begin{equation} \label{eq:eigenvalues_greater_than_1_condition}
\begin{aligned}
 \alpha_2 > 
 \frac{-b (T_A-T_B) \tilde{p}^* - (T_A - T_B)  + c (1-T_B)}{bT_A - b (T_A-T_B) \tilde{p}^*} = %\\
% \frac{-\frac{c(1-T_B) - b \alpha_1 T_A - (T_A - T_B)}{1-\alpha_1} - (T_A - T_B)  + c (1-T_B)}{bT_A - \frac{c(1-T_B) - b \alpha_1 T_A - (T_A - T_B)}{1-\alpha_1}} = \\
% \frac{\frac{-c(1-T_B) + b \alpha_1 T_A + (T_A - T_B)- (1-\alpha_1)(T_A - T_B)  + (1-\alpha_1)c (1-T_B)}{1-\alpha_1}}{\frac{(1-\alpha_1)bT_A - c(1-T_B) + b \alpha_1 T_A + (T_A - T_B)}{1-\alpha_1}} = \\
%  \frac{-c(1-T_B) + b \alpha_1 T_A + (T_A - T_B)- (1-\alpha_1)(T_A - T_B)  + (1-\alpha_1)c (1-T_B)}{(1-\alpha_1)bT_A - c(1-T_B) + b \alpha_1 T_A + (T_A - T_B)} = \\
%\frac{\alpha_1 b  T_A + \alpha_1 (T_A - T_B)  - \alpha_1 c (1-T_B)}{bT_A - c(1-T_B) + (T_A - T_B)} = 
\alpha_1 \;,
\end{aligned}
\end{equation}
which gives the following result.
\\

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{} \label{sec:appendixA}

@@

@@

%%% Table: payoff matrix
\bigskip
\begin{table}[h]
\centering
\caption{\textbf{Payoff matrix for prisoner's dilemma.}}
\begin{tabular}{lll}
\toprule
           & $\phi_2=A$ & $\phi_2=B$ \\ \cmidrule(r){1-3}
$\phi_1=A$ & $1+b-c$ & $1-c$ \\
$\phi_1=B$ & $1+b$   & $1$
\\ \bottomrule
\end{tabular}
\label{table:prisoner_payoff}
\caption*{
The fitness of phenotype $\phi_1$ when interacting with phenotype $\phi_2$. $A$ is a cooperative phenotype, $B$ is a defector phenotype, $b$ is the benefit gained by an individual interacting with a cooperator, and $c$ is the cost of cooperation. $0<b<c$.
}
\end{table}
\bigskip
%%%

@@

@@

% Fluctuations in spatial model
\begin{figure}[h]
  \centering
   \includegraphics[scale=0.75]{{timeSereis_b=1.3_TA=0.4_TB=0.435_c=0.02_globalSelection_normalizeByNumInteractions}.pdf}
   \caption{
    \textbf{Stable coexistence of both phenotypes in a structured population.}
    The frequency of cooperators (green) and defectors (red) in a structured finite population.
    Both phenotypes start at 50\% frequency.
    The figure shows that neither phenotype is fixed throughout the simulation, maintaining stochastic coexistence.
    Here, population size is 10,000 (100-by-100 grid), selection operates globally, 
    benefit and cost of cooperation $b=1.3$ and $c=0.02$, perfect vertical transmission $v=1$, horizontal transmission rates $T_A = 0.4 < T_B = 0.435$.
    }
    \label{fig:spatial_fluctuations}
  \end{figure}
  %TODO: edit this caption. DC

>>>>

We determined the equilibria of our model, i.e. the fixed points of \autoref{eq:nextgen_juveniles}, and their local stability.
The equilibria are $\hat p=0$, $\hat p=1$, and (see \autoref{eq:general_equilibrium_appendix})
\begin{equation} \label{eq:general_equilibrium}
  \hat{p}^* =  
  \frac{\alpha bvT_A - cv(1-T_B) + (T_A-T_B)}{\big[c(1-v) - b (1-\alpha v)\big] (T_A-T_B)} \;.
\end{equation}

To find conditions on the cost of cooperation, $c$, we define the following cost boundaries, $\gamma_1$ and $\gamma_2$, vertical transmission threshold, $\hat v$,
\begin{equation} \label{eq:cost_boundaries_v_threshold} 
\gamma_1 = \frac{b v \alpha T_A + (T_A - T_B)}{v(1-T_B)}, \quad
\gamma_2 = \frac{b v \alpha T_B + (1+b) (T_A - T_B)}{v(1-T_B) + (1-v)(T_A-T_B)}, \quad
\hat v = \frac{T_B - T_A}{1-T_A} \;.
\end{equation}

The full analysis is given in Appendix~\ref{sec:appendixB}.
The following result summarizes the possible outcomes.
\\

@@

@@

We now analyze the evolution of social association under perfect vertical transmission, $v=1$. 
We assume a population at a stable coexistence of the two phenotypes, cooperation $A$, and defection $B$, such that the frequency of $A$ among juveniles is $\hat{p}^*$ (\autoref{eq:general_equilibrium}).
Note that for stable coexistence to occur, we assume horizontal bias for defection, $T_A < T_B$, and an intermediate cost of cooperation, $\gamma_2 < c < \gamma_1$ (\autoref{eq:cost_boundaries_v_threshold})

The stable population mean fitness is $\bar{w}^* = 1 + \hat{p}^*(b-c)$, which is increasing in $\hat{p}^*$. 
In turn, $\hat{p}^*$ is also increasing in $\alpha$ (Appendix~\ref{sec:appendixC}).
Therefore, if social association increases, the population mean fitness also increases.
But can an allele that increases social association evolve? 
To answer this question, we extend our model to include a ``modifier locus''~\citep{Liberman1986,Liberman1988} that determines social association, but has no direct effect on fitness.
Assume that the modifier locus has two alleles, $M$ and $m$, which induce social associations $\alpha_1$ and $\alpha_2$, respectively.
Suppose that the population evolves to a stable equilibrium when only allele $M$ is present.
We check the local stability of this equilibrium to the invasion of the modifier allele $m$; this stablity is called ``external stability'', see \citet{Altenberg2017}.

The full analysis is given in Appendix~\ref{sec:appendixD}, and following result summarizes the outcomes of such invasions.
\\

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Local stability criterion} \label{sec:appendixA}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Equilibria and stability} \label{sec:appendixB}

We define $f(\hat{p}) = \bar{w}(\hat{p}' - \hat{p})$.
Then, using \emph{SymPy} \citep{Meurer2017}, a Python library for symbolic mathematics, this simplifies to
\begin{equation} \label{eq:general_case_polynomial}
\begin{aligned}
  f(\hat{p}) &= \bar{w}(\hat{p}'-\hat{p}) =
  \beta_1 \hat{p}^3 + \beta_2 \hat{p}^2 + \beta_3 \hat{p} \;,
\end{aligned}
\end{equation}
where 
\begin{equation} \label{eq:polynomial_coefficients}
\begin{aligned}
  \beta_1 &= \big[c(1-v) - b (1-\alpha v)\big] (T_A-T_B) \;, \\
  \beta_2 &= -\beta_1 -\beta_3 \;, \\
  \beta_3 &= \alpha bvT_A - cv(1-T_B) + (T_A-T_B) \;.
\end{aligned}
\end{equation}

In the general case where $T_A \neq T_B$, the coefficient $\beta_1$ is not necessarily zero, and $f(\hat{p})$ is a cubic polynomial.
Therefore, three equilibria may exist, two of which are
$\hat{p} = 0 $ and $\hat{p} = 1$, and the third is
\begin{equation} \label{eq:general_equilibrium_appendix}
  \hat{p}^* =  
  \frac{\beta_3}{\beta_1} =
  \frac{\alpha bvT_A - cv(1-T_B) + (T_A-T_B)}{\big[c(1-v) - b (1-\alpha v)\big] (T_A-T_B)} \;.
\end{equation}

This analysis can be summarized with the following conditions:
% in beta
\begin{enumerate}
\item \emph{Fixation of cooperation}: 
	if \emph{(i)} $T=T_A=T_B$ and $c < b\cdot \frac{\alpha T}{1-T}$; or
	if \emph{(ii)} $T_A>T_B$ and $0<\beta_3$; or 
	if \emph{(iii)} $T_A<T_B$ and $\beta_1<\beta_3$.
\item \emph{Fixation of the defection}: 
	if \emph{(iv)}  $T=T_A=T_B$ and $c > b\cdot \frac{\alpha T}{1-T}$; or 
	if \emph{(v)} $T_A>T_B$ and $\beta_3<\beta_1<0$; or 
	if \emph{(vi)} $T_A<T_B$ and $\beta_3<0$.
\item \emph{Coexistence of both phenotypes at $\hat{p}^*$}: 
	if \emph{(vii)} $T_A < T_B$ and $0<\beta_3<\beta_1$.
\item \emph{Fixation of either phenotype depending on initial frequency}:
	if \emph{(viii)}  $T_A>T_B$ and $\beta_1<\beta_3<0$.
\end{enumerate}

We now proceed to use the cost boundaries, $\gamma_1$ and $\gamma_2$, and the vertical transmission threshold, $\hat v$ (\autoref{eq:cost_boundaries_v_threshold}).
First, assume $T_A<T_B$.
$\beta_3<0$ requires $\gamma_1<c$.
For $\beta_3<\beta_1$ we need $c\big[v(1-T_B) + (1-v)(T_A-T_B)\big] > bv\alpha T_B + (1+b)(T_A-T_B)$.
Note that the expression in the square brackets is positive if and only if $v > \hat v$.
Thus, for $\beta_3<\beta_1$ we need $v > \hat v$ and $\gamma_2 < c$ or $v < \hat v$ and $c < \gamma_2$,
and for $0<\beta_3<\beta_1$ we need $v > \hat v$ and $\gamma_2 < c < \gamma_1$, or $v < \hat v$ and $c < \min(\gamma_1, \gamma_2)$. For $\beta_1<\beta_3$ we need $v > \hat v$ and $c<\gamma_2$ or $v < \hat v$ and $\gamma_2<c$.
However, some of these conditions cannot be met, since $v < \hat v$ implies $c<1<\gamma_2$.

This analysis is summarized in Result~\ref{result:vert_obli_hori}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Effect of social association on mean fitness} \label{sec:appendixC}

To determine the effect of increasing $\alpha$ on the stable population mean fitness, $\bar{w}^*=1+(b-c)\hat{p}^*$, we must analyze its effect on $\hat{p}^*$, 
\begin{equation} \label{eq:mean_fitness_stable_coexistence_derivative}
  \frac{\partial \hat{p}^*}{\partial \alpha} 
%  =\frac{-b^2T_A (1-\alpha) (T_A-T_B) + b(T_A-T_B)(c(1-T_B) - b \alpha T_A - (T_A-T_B))}{b^2 (1-\alpha)^2 (T_A-T_B)^2} 
%  = \frac{b^2T_A(T_B-T_A)-bc(1-T_B)(T_B-T_A) - b(T_B-T_A)^2}{b^2 (1-\alpha)^2 (T_A-T_B)^2} \;.
  = \frac{b T_A - c(1-T_B) + (T_A-T_B)}{b (1-\alpha)^2 (T_B-T_A)} \;.
\end{equation} 
Note that stable coexistence implies $c<\gamma_1$, and because $\alpha<1$, we have
\begin{equation}
c < \gamma_1 = \frac{b \alpha T_A + (T_A-T_B)}{1-T_B} < \frac{b T_A + (T_A-T_B)}{1-T_B}.
\end{equation} 
Therefore, the numerator in \autoref{eq:mean_fitness_stable_coexistence_derivative} is positive.
Since $T_A<T_B$, the denominator in \autoref{eq:mean_fitness_stable_coexistence_derivative} is also positive, and hence the derivative $\partial \hat{p}^* / \partial \alpha$ is positive.
Thus, the population mean fitness increases as social association $\alpha$ increases.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{External stability of $\tilde{p}^*$} \label{sec:appendixD}

We assume here that $v=1$, i.e. no oblique transmission, and therefore $\hat{p}=\tilde{p}$.
We define the frequencies of the pheno-genotypes $AM$, $BM$, $Am$, and $Bm$ as $\vec{\tilde{p}}= (\tilde{p}_1$, $\tilde{p}_2$, $\tilde{p}_3$, $\tilde{p}_4)$. 
The frequencies of the pheno-genotypes in the next generation are defined by the recursion system, 
\begin{equation} \label{eq:next_gen_p_1}
  \begin{aligned}
	% p1
  \bar{w}\tilde{p}_1' = 
  & \tilde{p}_1(\tilde{p}_1+\tilde{p}_3)(1+b-c)(1 - (1-\alpha_1)(\tilde{p}_2 + \tilde{p}_4)T_B) + \\
  & \tilde{p}_1(\tilde{p}_2 + \tilde{p}_4)(1-c)(1-\alpha_1T_B(\tilde{p}_1 + \tilde{p}_3) - T_B(\tilde{p}_2 + \tilde{p}_4)) + \\
  & \tilde{p}_2(\tilde{p}_1 + \tilde{p}_3)(1+b)T_A(\tilde{p}_1 + \tilde{p}_3 + \alpha_1(\tilde{p}_2 + \tilde{p}_4)) + \\
  & \tilde{p}_2(\tilde{p}_2 + \tilde{p}_4)(\tilde{p}_1 + \tilde{p}_3)(1-\alpha_1)T_A \;,
\\
	% p2
  \bar{w}\tilde{p}_2' = 
  & \tilde{p}_2(\tilde{p}_2+\tilde{p}_4)(1 - (1-\alpha_1)(\tilde{p}_1 + \tilde{p}_3)T_A) + \\
  & \tilde{p}_2(\tilde{p}_1 + \tilde{p}_3)(1+b)(1-\alpha_1 T_A(\tilde{p}_2 + \tilde{p}_4) - T_A(\tilde{p}_1 + \tilde{p}_3)) + \\
  & \tilde{p}_1(\tilde{p}_2 + \tilde{p}_4)(1-c)(\alpha_1 T_B + (1-\alpha_1)(\tilde{p}_2 + \tilde{p}_4)T_B) + \\
  & \tilde{p}_1(\tilde{p}_1 + \tilde{p}_3)(1+b-c)(1-\alpha_1)(\tilde{p}_2 + \tilde{p}_4)T_B \;, 
\\
	% p3
  \bar{w}\tilde{p}_3' =
  & \tilde{p}_3(\tilde{p}_1+\tilde{p}_3)(1+b-c)(1 - (1-\alpha_2)(\tilde{p}_2 + \tilde{p}_4)T_B) + \\
  & \tilde{p}_3(\tilde{p}_2 + \tilde{p}_4)(1-c)(1-\alpha_2T_B(\tilde{p}_1 + \tilde{p}_3) - T_B(\tilde{p}_2 + \tilde{p}_4)) + \\
  & \tilde{p}_4(\tilde{p}_1 + \tilde{p}_3)(1+b)T_A(\tilde{p}_1 + \tilde{p}_3 + \alpha_2(\tilde{p}_2 + \tilde{p}_4)) + \\
  & \tilde{p}_4(\tilde{p}_2 + \tilde{p}_4)(\tilde{p}_1 + \tilde{p}_3)(1-\alpha_2)T_A \;, 
\\
	% p4
  \bar{w}\tilde{p}_4' =
  & \tilde{p}_4(\tilde{p}_2 + \tilde{p}_4)(1 - (1-\alpha_2)(\tilde{p}_1 + \tilde{p}_3)T_A) + \\
  & \tilde{p}_4(\tilde{p}_1 + \tilde{p}_3)(1+b)(1-\alpha_2T_A(\tilde{p}_2 + \tilde{p}_4)-T_A(\tilde{p}_1 + \tilde{p}_3)) + \\
  & \tilde{p}_3(\tilde{p}_2 + \tilde{p}_4)(1-c)(\alpha_2T_B + (1-\alpha_2)(\tilde{p}_2 + \tilde{p}_4)T_B) + \\
  & \tilde{p}_3(\tilde{p}_1 + \tilde{p}_3)(1+b-c)(1-\alpha_2)(\tilde{p}_2 + \tilde{p}_4)T_B \;,
  \end{aligned}
\end{equation}
%which we can write as 
%\begin{equation}
%\bar{w} \vec{\tilde{p}} = F(\vec{\tilde{p}}),
%\end{equation}
where $\bar{w} = 1 + (b-c)(\tilde{p}_1 + \tilde{p}_3)$ is the population mean fitness.% and $\vec{\tilde{p}} = (\tilde{p}_1,\tilde{p}_2,\tilde{p}_3,\tilde{p}_4)$.

If both eigenvalues of $\mathbf{L}^*_{ex}$ are less than 1, then the equilibrium $\vec{\tilde{p}^*}$ is externally stable and allele $m$ cannot invade the population of allele $M$. 
These eigenvalues are the root of the characteristic polynomial $\lambda^2 -\lambda (X+Q) + XQ - YZ$; the leading eigenvalue $\lambda_1$ is the largest of them,
\begin{equation} \label{eq:eigenvalues}
  \lambda_1 = \frac{X+Q + \sqrt{(X+Q)^2 - 4(XQ-YZ)}}{2} \;.
\end{equation}
Using \emph{SymPy} \citep{Meurer2017}, a Python library for symbolic mathematics, we find that $\lambda_1 > 1$ if 
\begin{equation} \label{eq:eigenvalues_greater_than_1_condition}
\begin{aligned}
 \alpha_2 > 
 \frac{-b (T_A-T_B) \tilde{p}^* - (T_A - T_B)  + c (1-T_B)}{bT_A - b (T_A-T_B) \tilde{p}^*} = %\\
% \frac{-\frac{c(1-T_B) - b \alpha_1 T_A - (T_A - T_B)}{1-\alpha_1} - (T_A - T_B)  + c (1-T_B)}{bT_A - \frac{c(1-T_B) - b \alpha_1 T_A - (T_A - T_B)}{1-\alpha_1}} = \\
% \frac{\frac{-c(1-T_B) + b \alpha_1 T_A + (T_A - T_B)- (1-\alpha_1)(T_A - T_B)  + (1-\alpha_1)c (1-T_B)}{1-\alpha_1}}{\frac{(1-\alpha_1)bT_A - c(1-T_B) + b \alpha_1 T_A + (T_A - T_B)}{1-\alpha_1}} = \\
%  \frac{-c(1-T_B) + b \alpha_1 T_A + (T_A - T_B)- (1-\alpha_1)(T_A - T_B)  + (1-\alpha_1)c (1-T_B)}{(1-\alpha_1)bT_A - c(1-T_B) + b \alpha_1 T_A + (T_A - T_B)} = \\
%\frac{\alpha_1 b  T_A + \alpha_1 (T_A - T_B)  - \alpha_1 c (1-T_B)}{bT_A - c(1-T_B) + (T_A - T_B)} = 
\alpha_1 \;.
\end{aligned}
\end{equation}

@@

%%% Table: payoff matrix
\bigskip
\begin{table}[h]
\centering
\caption{\textbf{Payoff matrix for prisoner's dilemma.}}
\begin{tabular}{lll}
\toprule
           & $\phi_2=A$ & $\phi_2=B$ \\ \cmidrule(r){1-3}
$\phi_1=A$ & $1+b-c$ & $1-c$ \\
$\phi_1=B$ & $1+b$   & $1$
\\ \bottomrule
\end{tabular}
\label{table:prisoner_payoff}
\caption*{
The fitness of phenotype $\phi_1$ when interacting with phenotype $\phi_2$. $A$ is a cooperative phenotype, $B$ is a defector phenotype, $b$ is the benefit gained by an individual interacting with a cooperator, and $c$ is the cost of cooperation. $0<c<1$ and $c<b$.
}
\end{table}
\bigskip
%%%

@@

% Fluctuations in spatial model
\begin{figure}[h]
  \centering
   \includegraphics[scale=0.75]{{timeSereis_b=1.3_TA=0.4_TB=0.435_c=0.02_globalSelection_normalizeByNumInteractions}.pdf}
   \caption{
    \textbf{Stable coexistence of both phenotypes in a structured population.}
    The frequency of cooperators (green) and defectors (red) in a structured finite population.
    Both phenotypes start at 50\% frequency.
    The figure shows that neither phenotype is fixed throughout the simulation, maintaining stochastic coexistence.
    Here, population size is 10,000 (100-by-100 grid), selection operates globally, 
    benefit and cost of cooperation $b=1.3$ and $c=0.02$, perfect vertical transmission $v=1$, horizontal transmission rates $T_A = 0.4 < T_B = 0.435$.
    }
    \label{fig:spatial_fluctuations}
  \end{figure}

@@
